The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-09-03 floats/figures.tex
General reread and polish

\emph{See to-do notes on Figure \ref{fig:locmap}
    \irina[inline]{include all geographic names}
    \arjen[inline]{elevation scale white-end confused with glaciers}
    \arjen[inline]{remove Arctic circle and miles scale}
    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
and \ref{fig:ivolarea}
    \julien[inline]{Make the figure print-proof and highlight markers corresponding to the ``best'' runs.}
}

>>>>

\emph{See to-do notes on Figure \ref{fig:locmap}
    \irina{include all geographic names}
    \arjen{elevation scale white-end confused with glaciers}
    \arjen{remove Arctic circle and miles scale}
    \arjen{use CIS, LIS and IIS for ice sheet names}
    \arjen{use N, S, E and W instead of minus, add 75\degree N label}
and \ref{fig:ivolarea}
    \julien{Make the figure print-proof and highlight markers corresponding to the ``best'' runs.}
}

### 2013-09-03 sections/abstract.tex
General reread and polish

\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model, using temperature offsets from present-day climatologies from five different datasets. Monthly mean surface air temperature and monthly precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance using a positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. Best results were obtained using the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
\end{abstract}

>>>>

\begin{abstract}
We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), using temperature offsets from present-day climatologies from five different datasets. Monthly mean surface air temperature and monthly precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance using a positive degree-day model. Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing. We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska. The closest match between model output and the reconstructed ice margin was obtained using climate forcing from the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
\end{abstract}

### 2013-09-03 sections/climate.tex
General reread and polish

\julien{Qiong, could you help me with one or two sentences on why we smooth CFSR data?}
An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}). A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neigbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.

>>>>

An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}). A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}. Finally, input files for PISM were prepared using \citet{web:nc4py}.
\julien{Qiong, could you help me with one or two sentences on why we smooth CFSR data?}

### 2013-09-03 sections/concl.tex
General reread and polish

For the Cordilleran Ice Sheet, we achieve a best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}. The latter dataset is preferable in our case due to the lack of WorldClim data offshore.

One must keep in mind, however, that these results are limited to our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets). At LGM, the presence of ice sheets most probably significantly affected circulation patterns and distribution of rainfall. At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.

>>>>

For the Cordilleran Ice Sheet, we achieve a best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}. The latter dataset is preferable in our case due to the lack of WorldClim data offshore. NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.

One must keep in mind, however, that these results are limited to our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets). At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of rainfall. At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.

### 2013-09-03 sections/intro.tex
General reread and polish

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.
\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}

@@

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

>>>>

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.

@@

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-09-03 sections/model.tex
General reread and polish

SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$. We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
\needref
\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}

@@

Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}. \julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig. 1, not model input.} It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function. The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}. The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

>>>>

SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$. We apply a range of till friction angles between 10 and 30\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
\needref
\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}

@@

Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig. 1, not model input.} It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function. The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

### 2013-09-03 sections/results.tex
General reread and polish

Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}). Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).

>>>>

Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}). Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}), final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).
